### 2016-02-03 doc/thesis/intro.tex
Updates?

@@

\gls{ABC} is the name for a family of techniques which can be used to fit
complex models to observed data. We shall make this precise below, but to fully
describe and motivate \gls{ABC}, it is necessary to first make explicit exactly
what we mean by ``fitting'' a model.

A linear regression is a model of the relationship between some data points
$\vec{x} = (x_1, \ldots, x_n)$ and outcomes $\vec{y} = (y_1, \ldots, y_n)$.
This model assumes that the outcomes are linearly related to the data, modulo
some noise caused by measurement errors, environmental fluctuations, and so on.
In other words, there is a constant $b$ such that $y_i = bx_i + \varepsilon_i$,
where $\varepsilon_i$ is the error associated with the outcome $y_i$. If we
make the additional assumption that the errors are normally distributed, then
the model takes the form
\begin{align}
  y_i = bx_i + \N(0, \sigma^2).
  \label{eq:regression}
\end{align}
When we say that we want to fit this model to observed data, we mean that we
want to find particular values for $b$ and $\sigma$ which are in some way
optimal given $\vec{x}$ and $\vec{y}$. We shall make this precise below. Notice
that, if we fix $b$ and $\sigma$, then we can write down the probability density
of observing $y_i$ given $x_i$. If $f_\N(\cdot \mid \mu, \sigma)$ is the
\gls{pdf} of the normal distribution with mean $\mu$ and variance $\sigma^2$,
and the $y_i$ are all independent, then
\[
  f(\vec{y} \mid \vec{x}, b, \sigma) = 
  \prod_{i=1}^n f_\N(y_i - bx_i \mid 0, \sigma^2).
\]
This gives us a natural criterion for choosing the optimal parameters: we want
to pick $b$ and $\sigma$ which define a model where the probability density of
our observed data is as high as possible. When performing such an optimization,
the above density function is also called the \defn{likelihood} of the
parameters,
\[
  \L(b, \sigma \mid \vec{x}, \vec{y}) = f(\vec{y} \mid \vec{x}, b, \sigma).
\]
The particular $b$ and $\sigma$ which optimize $\L(b, \sigma)$ are called the
\gls{ML} estimates.

%Mathematical models can be grouped into three categories in terms of our
%ability to fit them to observed data. First, there is a limited set of models
%which can be fit explicitly by using calculus to minimize a loss function. The
%canonical example of this type of model is a linear regression. 
%Notice that, if we fix $b$, then since $y_i$ and $x_i$ are both given,
%$\varepsilon_i$ is also fixed. The process of fitting a linear model means
%choosing the ``best'' $b$. For least squares, this optimal $b$ is defined as
%the one which minimizes the sum of the squared $\varepsilon_i$.

Now that we have defined what it means to fit a least-squares linear
regression, there remains the question of how to go about finding this optimal
$b$ value. Recall that the errors $\varepsilon_i$ are fixed for a particular
$b$, and our objective is to minimize the loss function, which is the sum of
the squared errors. Naively, we could simply try many different $b$ values,
calculate the loss function for each, and choose the $b$ which gives the
smallest value. This approach, though basic, falls under the umbrella of
\defn{numerical optimization}, which is the most common way the majority of
models are fit. Of course, there is a whole field devoted to more sophisticated
methods for exploring the parameter space, but they all boil down to the basic
method of trying different values and choosing the one which minimizes the
loss.

In the case of least-squares linear regression, there is another method we
could use. Looking at equation~\ref{eq:regression} for the model, it is easy to
rearrange terms to solve for $\varepsilon_i$ in terms of the other variables:
we get $\varepsilon_i = y_i - bx_i$. In fact, we can write down the entire loss
function, which is the sum of the squared $\varepsilon_i$ in terms of $b$:
\[
  \sum \varepsilon_i^2 = \sum (y_i - bx_i)^2,
\]
where the sums are over all $i$ from 1 to $n$, the number of points in the data
set. We will not describe the procedure here, but it is possible to explicitly
minimize the right-hand side by finding the point where the derivative with
respect to $b$ is zero. However, the method of directly minimizing the loss
function with calculus is only applicable to a very narrow class of simple
models. For the majority, the loss function is too complicated to set the
derivative with respect to the parameters to zero and solve.

Because of the limited space of models which permit exact solutions, numerical
optimization of a loss function is the most frequently used method to fit a
model.

>>>>

An alternative to summary statistics are \defn{distance measures} on trees.
Rather than assigning a numerical value to each tree individually, a distance
measure assocates each pair of trees with a number, indicating how different
the trees are from each other. Distance measures allow us to identify groups of
related phylogenies, for example, local epidemics which are undergoing a
similar pattern of expansion.

@@

\Gls{ABC} is a statistical method designed to fit complex models, which cannot
be fit using more conventional methods, to observed
\comment{these are review articles, Beaumont is specific to popgen, the others
are general}
data~\autocite{marin2012approximate, sunnaker2013approximate,
beaumont2010approximate} We shall make this precise below, but to fully
describe and motivate \gls{ABC}, it is necessary to first explain exactly what
we mean by ``fitting'' a model. We will illustrate this concept with one of the
simplest and most ubiquitous models: linear regression.

A linear regression is a model of the relationship between some data $\vec{x} =
(x_1, \ldots, x_n)$ and outcomes $\vec{y} = (y_1, \ldots, y_n)$. This model
assumes that the outcomes are linearly related to the data, modulo some noise
introduced by, say, measurement errors and environmental fluctuations. It other
words, there is a constant $\beta$ such that $y_i = \beta x_i + \varepsilon_i$,
where $\varepsilon_i$ is the error associated with the outcome $y_i$. If we
make the additional assumption that the errors are normally distributed, then
the model takes the form
\[
  y_i = \beta x_i + \N(0, \sigma^2),
\]
where $\N(0, \sigma^2)$ indicates a normally distributed random variable with
mean 0 and variance $\sigma$. We will denote the \gls{pdf} of this random
variable by $f_\N(\cdot \mid \mu, \sigma)$. In this formulation, the
coefficient $\beta$ and the error variance $\sigma$ are the two parameters of
the model. If the $y_i$ are all independent, then for particular choics of
$\beta$ and $\sigma$, we can write down the probability density of observing
$\vec{y}$ given $\vec{x}$.
\[
  f(\vec{y} \mid \vec{x}, \beta, \sigma) = 
  \prod_{i=1}^n f_\N(y_i - \beta x_i \mid 0, \sigma^2).
\]
The higher the value of this probability density, the more likely the
observations $\vec{y}$ are given $\vec{x}$ under the model. This gives us a
natural criterion for choosing the parameters: we want to pick $\beta$ and
$\sigma$ which define a model where the probability of our observed data is as
high as possible. When performing such an optimization for fixed data, the
density function just written is also called the \defn{likelihood} of the
parameters,
\[
  \L(\beta, \sigma \mid \vec{x}, \vec{y}) = f(\vec{y} \mid \vec{x}, \beta, \sigma).
\]
The particular $\beta$ and $\sigma$ which optimize $\L(\beta, \sigma \mid
\vec{x}, \vec{y})$ are called the \gls{ML} estimates.

\Gls{ML} inference makes use only of the data at hand, in this case
$\vec{x}$ and $\vec{y}$, to estimate the parameters $\beta$ and $\sigma$.
However, it is frequently the case that the investigator has some additional
\defn{prior} information or belief about what $\beta$ and $\sigma$ are likely
to be. For example, the instrument used to measure the $\vec{y}$ may have a
known margin of error, or the sign of the slope may be obvious from previous
experiments. A Bayesian approach to fitting this model would make use of this
information by codifying the investigator's beliefs as a \defn{prior
distribution}, denoted $f(\beta, \sigma)$. The optimal parameters are then
those which maximize the product of the prior and the likelihood,
\[
  f(\vec{y} \mid \vec{x}, \beta, \sigma) f(\beta, \sigma).
\]
Bayes' theorem tells us that this product is proportional to the
\defn{posterior distribution} $f(\beta, \sigma \mid \vec{y}, \vec{x})$. The
parameters which optimize the posterior, and hence also optimize the above
product, are called the \gls{MAP} estimates.

Now that we have defined what it means to fit a linear regression, there
remains the question of how to go about finding this optimal $\beta$ value.
Na\"{i}vely, we could simply try many different $\beta$ and $\sigma$ values,
calculate the likelihood of each, and choose those which yield the highest
value. This approach, though basic, falls under the umbrella of \defn{numerical
optimization} of the likelihood function. Of course, there is a whole field
devoted to more sophisticated methods for exploring the parameter space, but
they all boil down to the basic idea of trying out different values and
choosing those which give the highest likelihood.

In the case of linear regression, there is another method we could use. We know
from calculus that the maximum of $\L$ occurs either on the boundary of its
domain, or at a point where its partial derivatives with respect to $\beta$ and
$\sigma$ are both zero. Though we will not show it here, it turns out to be
straightforward to find the $\beta$ and $\sigma$ values where the partial
derivatives are zero. In fact, these values co-incide with the least-squares
estimates of $\beta$ and $\sigma$, which minimize the sum of the squared $x_i -
\beta y_i$. However, the method of directly minimizing the loss function with
calculus is only applicable to a narrow class of simple models. For the
majority, the likelihood function is too complicated to set the partial
derivatives with respect to the parameters to zero and solve.
